---
title: "Lactate Analysis Performed on the Tarragona Database"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'indexLactateTarragona.html'))})
author: "Giovana Gavidia"
date: "May 10, 2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
source(file='mainFunctionsMimics.R')
```

## Read Data
```{r message=FALSE, warning=FALSE}
PatInfo <- read.csv("~/Documents/Projects/DATATHON/MIMIC/data/Tarragona/PatInfoApril2019.csv", stringsAsFactors=FALSE)
# info demografica de los pacientes a 1 linia por paciente. Respecto al que ya tenias hay una columna más que se llama VMI, que es un 1 si llevaron ventilación mecánica invasiva el primer dia de estancia y un NA si no la llevaron. Aquí está el APACHEII para hacer los grupos por deciles.
# 

# 
# AllVarsApril2019 <- Tienes todos los valores de las variables de interés en el primer dia de estancia resumidos (minimo, maximo, media etc...). Aquí está el lactato.

FirstDayVars <- read.csv("~/Documents/Projects/DATATHON/MIMIC/data/Tarragona/AllVarsApril2019.csv", stringsAsFactors=FALSE)

# 
# DrugsApril2019 <- Tienes las drogas vasoactivas que se dieron el primer dia de estancia y sus dosis. Dopamine igual que el MIT, Epinephrine y Norepinephrine son Adrenaline y Noradrenaline en nuestro caso, son las mismas drogas con diferente nombre, y finalmente Phenylephrine y vasopressin no usamos... Te las doy por separado para que puedas hacer analisis por separado, pero si quieres hacer analisis de dosis de drogas vasoactivas total tendrás que sumar las dosis de todas las que haya llevado un mismo paciente, y para tener el numero pues contar cuantas diferentes han recibido.
Vasopressors_FirstDay <- read.csv("~/Documents/Projects/DATATHON/MIMIC/data/Tarragona/DrugsApril2019.csv", stringsAsFactors=FALSE)

```

# Preprocessing

```{r message=FALSE, warning=FALSE, echo=FALSE}

# Replace ',' by .
FirstDayVars<-data.frame(FirstDayVars[,c('PatientID','VariableID','VarName')], apply(FirstDayVars[,c('FirstValue','LastValue','MaxValue','MinValue','AvgValue','StdvValue', 'MedianValue','IQR1Value','IQR3Value')],2, covertComma2Period) ) 
```

## Initial Lactate vs. APACHE Analysis
```{r message=FALSE, warning=FALSE, echo=FALSE, results='asis',fig.height = 7, fig.width = 10, fig.align = "center"}

# Focus only the Max values computed to Lactate
d.lactate<-subset(FirstDayVars, VarName %in% "Lactate")

# Several participants have more that one lactate value, so only the biggest value is considered

library(reshape2)
d.lactate.max <- dcast(d.lactate[,c('PatientID','VarName','MaxValue')], PatientID ~ VarName, value.var="MaxValue",fun=max)

#---- classify APACHE using deciles

# quantile(PatInfo$APACHEII, prob = seq(0, 1, length = 11), type = 5)

PatInfo$APACHEII_dec<-cut( PatInfo$APACHEII, quantile(PatInfo$APACHEII, prob = seq(0, 1, length = 11), type = 5),labels=seq(1,10,1),include.lowest = TRUE )

d.lactate.max<-merge(d.lactate.max,PatInfo[,c('PatientID','APACHEII','APACHEII_dec')],by='PatientID',all.x = TRUE)

print("15 patients who have lactate values do not have APACHE info, so, they were removed")
d.lactate.max<-subset(d.lactate.max,!is.na(APACHEII_dec))

library(psych)
res<-describeBy(group='APACHEII_dec',  x = d.lactate.max[,c('APACHEII_dec','Lactate')],mat=TRUE)
res<-res[11:20,c('group1','n','mean','sd','median','trimmed','mad','min','max','range','skew','kurtosis','se')]
rownames(res)<-NULL

colnames(res)[which(colnames(res) %in% 'group1')]<-'APACHEgroup'

# library(xtable)
library(kableExtra)
# xtb<-xtable(res, caption = 'Summary of the Max Lactate values according to the APACHE categories', label = 'Tab:mRMR')
# print(xtb, type = "html")

#kable(res,caption = 'Summary of the Max Lactate values according to the APACHE categories')
res %>%
  kable() %>%
  kable_styling()


```

* Distribution of the Lactate valeus according to the APACHE categories

```{r message=FALSE, warning=FALSE, echo=FALSE, results='asis',fig.height = 7, fig.width = 10, fig.align = "center"}
# d<-d.lactate.max[,c('APACHEII_dec','Lactate')]
# plotStatisticsWhole(d,plot_title='Exploratory Analysis of the Continuous Variables by APACHE group',source='Lactate',sizeFont=5,groupname='APACHEII_dec')


# Plot the distribution of lactate according to each APACHE group
library(ggplot2)
library(tidyverse)
library(hrbrthemes)

options(knitr.table.format = "html")


# file<-paste('images/distribution_APACHEgroups.png',sep='')
#   #postscript(file=file, paper="special", width=6, height=6)
#   png(file=file,width=30,height=15,units="cm",res=150)
  
print(ggplot(d.lactate.max, aes(x=Lactate, fill=APACHEII_dec)) + geom_density(alpha=.3)+
  xlab("APACHE groups") + 
    ylab("Density") +       
    theme_bw() +
    theme(plot.title = element_text(size = 14, family = "Helvetica", face = "bold"),
    text = element_text(size = 14, family = "Helvetica"),
    axis.title = element_text(face="bold"),
    axis.text.x=element_text(size = 14)))
# dev.off()

# file<-paste('images/numberSamples_APACHEgroups.png',sep='')
#   #postscript(file=file, paper="special", width=6, height=6)
#   png(file=file,width=30,height=15,units="cm",res=150)

```

* Figure below show that the 5th decile has the largest number of patients.

```{r message=FALSE, warning=FALSE, echo=FALSE, results='asis',fig.height = 7, fig.width = 10, fig.align = "center"}
  
res %>%
  filter(!is.na(n)) %>%
  arrange(n) %>%
  mutate(APACHEgroup=factor(APACHEgroup, APACHEgroup)) %>%
  ggplot( aes(x=APACHEgroup, y=n) ) +
    geom_segment( aes(x=APACHEgroup ,xend=APACHEgroup, y=0, yend=n), color="grey") +
    geom_point(size=3, color="#69b3a2") +
    coord_flip() +
    theme_ipsum() +
    theme(
      panel.grid.minor.y = element_blank(),
      panel.grid.major.y = element_blank(),
      legend.position="none"
    ) +
    xlab("APACHE groups") + 
    ylab("N")
   
  # dev.off()
```

## Exploration of ICU Tratments Differences

We want to visualize the total dose of vasopressor on the first day in the ICU for each drug listed here:

* Dopamine
* Adrenaline (in MIMICS, it is equivalent to Epinephrine)
* Noradrenaline (in MIMICS, it is equivalent to Norepinephrine)

 In Tarragona, thet do not use Phenylephrine and  Vasopressin


Figure below shows the missingness pattern in data according to Gender, VMI and drugs. Percentages represent the amount of missing observation for each variables. Missing samples are highlighted in black. In addition, the graph also provide information on the overall percentage of missing values overall (in the legend). Note that almost 29% of patients were not assisted with a VMI (Invasive Mechanical Ventilation).



```{r message=FALSE, warning=FALSE, echo=FALSE, results='asis', fig.height = 7, fig.width = 10, fig.align = "center"}

# DrugsApril2019 <- Tienes las drogas vasoactivas que se dieron el primer dia de estancia y sus dosis. Dopamine igual que el MIT, Epinephrine y Norepinephrine son Adrenaline y Noradrenaline en nuestro caso, son las mismas drogas con diferente nombre, y finalmente Phenylephrine y vasopressin no usamos... Te las doy por separado para que puedas hacer analisis por separado, pero si quieres hacer analisis de dosis de drogas vasoactivas total tendrás que sumar las dosis de todas las que haya llevado un mismo paciente, y para tener el numero pues contar cuantas diferentes han recibido.
Vasopressors_FirstDay <- read.csv("~/Documents/Projects/DATATHON/MIMIC/data/Tarragona/DrugsApril2019.csv", stringsAsFactors=FALSE)

d.Vasopressors<-merge(Vasopressors_FirstDay, PatInfo[,c("PatientID", "APACHEII","Gender", "VMI","APACHEII_dec")], by='PatientID', all.x = TRUE)
d.Vasopressors$Dose<-covertComma2Period(d.Vasopressors$Dose)

# reshape long 2 wide

library(reshape2)

dw.Vasopressors <- dcast(d.Vasopressors, PatientID + Gender + VMI + APACHEII_dec ~ Drug, value.var="Dose")

library(naniar)

print(vis_miss(dw.Vasopressors ))



```

* Following graph shows the usage of drugs according to the APACHE groups. Most of the patients were treated with Noradrenaline, where most of the patients belong to the APACHE Group 10
```{r message=FALSE, warning=FALSE, echo=FALSE, results='asis', fig.height = 7, fig.width = 10, fig.align = "center"}
library(ggplot2)
# plotStatisticsWhole(d=dw.Vasopressors,plot_title='',source='Patient',sizeFont=3.5,groupname='APACHEII_dec')

g <- ggplot(d.Vasopressors, aes(Drug))+
   geom_bar(aes(fill = APACHEII_dec), position = position_stack(reverse = TRUE)) +
 coord_flip() +
  theme_bw() +
    theme(plot.title = element_text(size = 14, family = "Helvetica", face = "bold"),
    text = element_text(size = 14, family = "Helvetica"),
    axis.title = element_text(face="bold"),
    axis.text.x=element_text(size = 14))
 #theme(legend.position = "top")

print(g)

```

## Next Steps: Obtain some plots considering normal and non-normal values of Lactate.